The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2025-04-22 Paper.qmd
Minor paper edits.

To gain support, children use signals to communicate their needs and wants to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has rarely been investigated in a single study of children of all ages. To fill this gap, we collected mother and other primary caregiver reports of child signaling from `{r} n_distinct(householdIDs)` families with `{r} nrow(utila_df)` children on Utila, a small island off the coast of Honduras. In exploratory analyses, we found that child signaling was common in both sexes and across all ages, although it decreased with age and neighborhood quality and increased with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were sad more frequently were perceived as needier within the household and were more likely to receive investment. Caregivers were less likely to respond positively in situations of family conflict or child transgressions, and more likely for injuries and illnesses. Our results suggest that evolutionary theories of signaling can help explain patterns of child sadness, crying, and temper tantrums.

@@

Although infant signals of need, particularly infant cries, have been extensively studied in many populations [@barr1991; @fouts2005; @konner1972; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, how and when older children signal need has rarely been investigated [for one exception, see @syme2023]. The aim of this study was to help fill these gaps.

@@

Humans are a uniquely interdependent species among the living primates [@tomasello2012] in that both sex's fitness prospects depend greatly on a network of close kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from mothers and fathers [@kramer2018]. However, grandparents, aunts, uncles, and non-relatives also provide significant investment in children in many societies [@kramer2018; @sear2008], which was likely favored due to a combination of inclusive fitness benefits, reciprocity, and reputation management [@gurven2000; @schniter2023]. Children also often serve as alloparents, helping to raise their younger siblings via direct care, food collection and preparation, and other family tasks [@hrdy2009; @meehan2009; @sear2008] despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

Sadness is often not accompanied by anger, suggesting little conflict with caregivers, and its costs are low when not prolonged. Temper tantrums, in constrast, are closely linked to anger [@potegal2003], suggesting conflict with caregivers, and have high costs even when the tantrum is brief. Thus, unlike sadness, temper tantrums might be useful when there are substantial and immediate gains to be had from a credible signal of need. The costs of temper tantrums could be energetic, or could come from the risk or implication of self-harm [@lummaa1998]. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, behaviors that have been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors could credibly signal need if the benefits of care for those who are truly in need outweigh the signal costs, whereas for those who are not in need, the benefits of care do not outweigh the signal costs, thus deterring deceptive signaling [@hagen2008].

@@

This study aimed to establish the natural history of child signaling in Utila, Honduras, by measuring various indices of adversity, children's conflicts with caregivers, the frequencies of common child signals of need, and various types of caregiver responses. From these, we used signaling theory to explore 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how caregivers respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses in two subpopulations that face different levels of social and economic adversity.

@@

One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1954], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the potential social costs of crying or otherwise expressing sadness [@glasberg1981; @glasberg1982; @zeman1996] against the potential increase in social support, which is consistent with our result that the age-related decline in signaling was steepest for crying. The age-related decline in signaling was also generally steeper for males than females [].

[new additions above too repetive?]

@@

Although this study was broader than most in that it examined sadness, crying, and temper tantrums across much of childhood and adolescence, it does not cover all the ways children might signal need. Pained facial and bodily expressions [@vigil2014; @craig2009; @cano2010; @schiavenato2010], self-harm outside of temper tantrums [@hagen2008; @nock2008], and suicidality [@syme2016] have all been suggested to credibly signal need in times of conflicts of interest and may be viable strategies for children, at least at certain ages. Running away has received similar treatment, especially in instances when it comes with substantial risk [@shostak1981] or is seen as a suicide-threat [@fischer1950; @syme2023]. For this reason, we also collected data on running away. However, mothers on Utila reported that running away on the island often involved children visiting neighbors or aunts and was much less risky than it would be on mainland Honduras, with a few mothers mentioning in focus groups that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean and the presence of dogs, motorcycles, golf carts, and tuk-tuks (auto rickshaws) on the narrow roads typical of much of populated parts of the island. Due to uncertainty on how to conceptualize the risks involved, we did not include running away in our analyses.

>>>>

To gain support, children use signals to communicate their needs and wants to parents. Infant signals of need, particularly infant cries, have been extensively studied in diverse populations. However, the full range of potential child signals of need, which extend beyond cries, has rarely been investigated in a single study of children of all ages. To help fill this gap, we collected mother and other primary caregiver reports of three common types of child signaling from `{r} n_distinct(householdIDs)` families with `{r} nrow(utila_df)` children on Utila, a small island off the coast of Honduras. In exploratory analyses, we found that child signaling was common in both sexes and across all ages, although it decreased with age and neighborhood quality and increased with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were sad more frequently were perceived as needier within the household and were more likely to receive investment. Caregivers were less likely to respond positively in situations of family conflict or child transgressions, and more likely for injuries and illnesses. Our results suggest that evolutionary theories of signaling can help explain patterns of child sadness, crying, and temper tantrums.

@@

Although infant signals of need, particularly infant cries, have been extensively studied in many populations [@barr1991; @fouts2005; @konner1972; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, how and when older children signal need has rarely been investigated [for one exception, see @syme2023]. The aim of this study was to help fill these gaps by investigating three common types of child signals.

@@

Humans are a uniquely interdependent species among the living primates [@tomasello2012] in that both sexes' fitness prospects depend greatly on a network of close kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from mothers and fathers [@kramer2018]. However, grandparents, aunts, uncles, and non-relatives also provide significant investment in children in many societies [@kramer2018; @sear2008], which was likely favored due to a combination of inclusive fitness benefits, reciprocity, and reputation management [@gurven2000; @schniter2023]. Children also often serve as alloparents, helping to raise their younger siblings via direct care, food collection and preparation, and other family tasks [@hrdy2009; @meehan2009; @sear2008] despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

Temper tantrums are closely linked to anger [@potegal2003], suggesting conflict with caregivers, and have high costs even when the tantrum is brief. Thus, temper tantrums might be useful when there are substantial and immediate gains to be had from a credible signal of need. This distinguishes them from sadness, and even crying, which are often unrelated to conflicts and have low costs when short-lived.

The costs of temper tantrums could be energetic, or could come from the risk or implication of self-harm [@lummaa1998]. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, behaviors that have been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors could credibly signal need if the benefits of care for those who are truly in need outweigh the signal costs, whereas for those who are not in need, the benefits of care do not outweigh the signal costs, thus deterring deceptive signaling [@hagen2008].

@@

This study aimed to establish the natural history of child signaling in Utila, Honduras, by measuring various indices of adversity, children's conflicts with caregivers, the frequencies of three common child signals of need, and various types of caregiver responses. From these, we used signaling theory to explore 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how caregivers respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses in two subpopulations that face different levels of social and economic adversity.

@@

One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1954], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the potential social costs of crying or otherwise expressing sadness [@glasberg1981; @glasberg1982; @zeman1996] against the potential increase in social support, which is consistent with our result that the age-related decline in signaling was steepest for crying. The age-related decline in signaling was also generally steeper for males than females [@sfig-signaltype].

@@

Although this study was broader than most in that it examined sadness, crying, and temper tantrums across much of childhood and adolescence, it did not cover all the ways that caregivers might infer need, such as cues of injury. It also did not investigate all potential signals of need. Pained facial and bodily expressions [@vigil2014; @craig2009; @cano2010; @schiavenato2010], self-harm outside of temper tantrums [@hagen2008; @nock2008], and suicidality [@syme2016] have all been suggested to credibly signal need in times of conflicts of interest and may be viable strategies for children, at least at certain ages. Running away has received similar treatment, especially in instances when it comes with substantial risk [@shostak1981] or is seen as a suicide-threat [@fischer1950; @syme2023]. For this reason, we also collected data on running away. However, mothers on Utila reported that running away on the island often involved children visiting neighbors or aunts and was much less risky than it would be on mainland Honduras, with a few mothers mentioning in focus groups that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean and the presence of dogs, motorcycles, golf carts, and tuk-tuks (auto rickshaws) on the narrow roads typical of much of populated parts of the island. Due to uncertainty on how to conceptualize the risks involved, we did not include running away in our analyses.
